---
title: "Birth Weights"
author: "Oscar Briones Ramirez"
date: '2023-01-30'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/oscar/Desktop/Winter2023/STAT469/Birthweights")

library(tidyverse)
library(GGally)
library(car)
library(MASS)
library(lmtest)
library(multcomp)

birth_weights <- read_table("BirthWeights.txt")
```

## EDA

**1. Scatterplot of BirthWeight by Mage**

```{r scatterplot, echo = TRUE}
#1. Scatterplot BirthWeight by Mage
ggplot(data=birth_weights, mapping=aes(x=Mage, y=BirthWeight)) + geom_point()+geom_smooth(se=FALSE)

```

**2. Side-by-side boxplots of BirthWeight for each category in Race **

```{r boxplot, echo = TRUE}
#2. boxplot
ggplot(data=birth_weights, mapping=aes(x=Race, y=BirthWeight)) + geom_boxplot()
```

**3. A scatterplot of BirthWeight by Gage where the dots are colored according to Gen **

```{r scatter2, echo = TRUE}

#3. Scatterplot BirthWeight by Gage where the dots are colored according to Gen
ggplot(data=birth_weights, mapping=aes(x=Gage, y=BirthWeight, color = Gen)) + geom_point()
```

**4. The correlation between BirthWeight and Mage. **

```{r cor, echo = TRUE}
#4. The correlation between BirthWeight and Mage.
cor(birth_weights$BirthWeight, birth_weights$Mage)
```

**5. A pairs plot of all the variables in the BirthWeight dataset. **

```{r pairs, echo = TRUE}
#5. Pairs plots

ggpairs(birth_weights)
```

## Fitting a Linear Model

```{r lm, echo = FALSE}
birth.lm <- lm(formula=BirthWeight~., data=birth_weights)

X <- model.matrix(object=BirthWeight~., data=birth_weights)
y <- birth_weights$BirthWeight
```

**1. Without the use of lm() calculate β̂ and s2. Verify your answer using lm().**
```{r Bhat, echo = TRUE}
Bhat <- solve((t(X)%*%X))%*%t(X)%*%y
coef(birth.lm)
Bhat

S2 <- (t(y-(X%*%Bhat))%*%(y-(X%*%Bhat)))/(832-4-1)
sigma(birth.lm)
sqrt(S2)
```


**2. Without the use of lm() calculate the fitted values Xβ̂ . Verify your calculations by pulling off the fitted values from an lm() object. **

```{r fitted, echo = TRUE}
fitvals <- X%*%Bhat
head(fitted(birth.lm))
head(fitvals)
```

**3. Without the use of lm() calculate the residuals y−Xβ̂  Verify your calculations by pulling off the residuals from an lm() object. **

```{r res, echo = TRUE}
resids <- y-X%*%Bhat
head(resid(birth.lm))
head(resids)
```

**4. Identify your model R2 from the summary() output. **
```{r rsq, echo = TRUE}
summary(birth.lm)$r.squared

```

## Checking Assumptions

**1. Construct added variable plots and assess if the linearity assumption is OK for this data. **

```{r av, echo = TRUE}
avPlots(birth.lm, ask=FALSE)

```

**2. Construct a histogram of the standardized residuals and run a KS-test to see if the normality assumption is OK for this data. **

```{r stdres, echo = TRUE}
ggplot() + geom_histogram(mapping=aes(x=stdres(birth.lm)))

ks.test(stdres(birth.lm), "pnorm")
```

**3. Draw a scatterplot of the fitted values vs. standardized residuals and run a BP-test to see if the equal variance assumption is OK for this data. **

```{r fittedvsres, echo = TRUE}
ggplot(mapping=aes(x=fitted(birth.lm), y=resid(birth.lm))) + geom_point()

bptest(birth.lm)
```

## Predictions


**1. Without using predict.lm(), calculate your point prediction of the birth weight for a baby with Mage=26,Gage=37, Race="hisp" and Gen="Female" using the formula ŷ new=xnewβ̂ where β̂ is the maximum likelihood estimate that you calculated above. Confirm that this is what predict.lm() is doing to get the point prediction.**

```{r pred1, echo = FALSE}
new.x = data.frame(Mage=26, Gage=37, Racehisp=1, Raceother=0, Racewhite=0, Genmale=0)
```

```{r pred, echo = TRUE}
newx <- data.frame(Intercept = 1, Mage=26, Gage=37, Racehisp=1, Raceother=0, Racewhite=0, GenMale=0)
ynew <- newx*Bhat
rowSums(ynew)

new.x = data.frame(Mage=26, Gage=37, Race="hisp", Gen="Female")
predict.lm(birth.lm, newdata=new.x, interval="prediction", level=0.99)
```

**2. Using predict.lm(), get a prediction of the birth weight for a baby with Mage=26, Gage=37, Race="hisp" and Gen="Female" and an associated 99% prediction interval.**

```{r pred2, echo = TRUE}
new.x = data.frame(Mage=26, Gage=37, Race="hisp", Gen="Female")
predict.lm(birth.lm, newdata=new.x, interval="prediction", level=0.99)

```

## Cross Validation

**1. Adjust the above code to run 100 Monte Carlo cross validations and plot histograms (or density plots) of the bias, RPMSE, coverage and width.**

```{r cv, echo = TRUE}

n.cv <- 100 #Number of CV studies to run
n.test <- 20  #Number of observations in a test set
rpmse <- rep(x=NA, times=n.cv)
bias <- rep(x=NA, times=n.cv)
wid <- rep(x=NA, times=n.cv)
cvg <- rep(x=NA, times=n.cv)
for(cv in 1:n.cv){
  ## Select test observations
  test.obs <- sample(x=1:n.cv, size=n.test)
  
  ## Split into test and training sets
  test.set <- birth_weights[test.obs,]
  train.set <- birth_weights[-test.obs,]
  
  ## Fit a lm() using the training data
  train.lm <- lm(formula=, data=train.set)
  
  ## Generate predictions for the test set
  my.preds <- predict.lm(train.lm, newdata=test.set, interval="prediction")
  
  ## Calculate bias
  bias[cv] <- mean(my.preds[,'fit']-test.set[['BirthWeight']])
  
  ## Calculate RPMSE
  rpmse[cv] <- (test.set[['BirthWeight']]-my.preds[,'fit'])^2 %>% mean() %>% sqrt()
  
  ## Calculate Coverage
  cvg[cv] <- ((test.set[['BirthWeight']] > my.preds[,'lwr']) & (test.set[['BirthWeight']] < my.preds[,'upr'])) %>% mean()
  
  ## Calculate Width
  wid[cv] <- (my.preds[,'upr'] - my.preds[,'lwr']) %>% mean()
  
}

ggplot() + geom_histogram(mapping=aes(x=bias))
ggplot() + geom_histogram(mapping=aes(x=rpmse))
ggplot() + geom_histogram(mapping=aes(x=cvg))
ggplot() + geom_histogram(mapping=aes(x=wid))

```

## Hypothesis Testing and Confidence Intervals

```{r hip, echo = FALSE}

reduced.lm <- lm(formula=BirthWeight~.-Mage-Gage, data=birth_weights)
full.lm <- lm(formula=BirthWeight~., data=birth_weights)

```

**1. Using lm() construct the t−statistic and p-value for the test H0:βMage=0.**

```{r hipt, echo = TRUE}
mage.lm <- lm(formula=BirthWeight~Mage, data=birth_weights)
summary(mage.lm)
```

**2. Using confint() and lm(), build a 90% confidence interval for βMage.**

```{r confint, echo = TRUE}
confint(mage.lm,level=.90)
```

**3. Using anova(), conduct a Ftest that race has no effect on birth weight (note: this answers primary research question #2).**

```{r ftest, echo = TRUE}
race.lm <- lm(formula=BirthWeight~Race, data=birth_weights)
anova(full.lm, race.lm)
```


**4. Using glht(), conduct a ttest and 94% confidence interval for the difference in average birth weight of babies born with explanatory variables**
```{r four, echo = TRUE}
at <- t((c(1, 24, 40, 0, 0, 1, 1)-c(1, 34, 33, 0, 0, 1, 1)))

my.test <- glht(full.lm, linfct=at, alternative="two.sided")
summary(my.test)

confint(my.test,level=.94)
```

